Supporting Information for\\ ``A weakly structured stem for human origins in Africa''

### 2023-01-17 supplement.tex
Update supplement.tex
add fixed params citations, format consistency

Our models also included recent events to account for known migrations,
admixtures, and growths and declines in effective population sizes. Many of
these parameters were fixed based on previous historical, genetic, or
anthropological research, namely
\begin{itemize}
    \item Western African vs. Eastern African divergence: 60ka \citep{Pagani2015-pz}
%Add Niang et al. "The Middle Stone Age occupations of Tiemassa, coastal West Africa, between 62 and 25 thousand years ago" Journal of Archaeological Science, 2020 as citation
%Add Chevrier et al. "Between continuity and discontinuity: An overview of the West African Paleolithic over the last 200,000 years". Quaternary International, 2018 as citation
    \item Eastern Africa vs. British divergence [i.e. OOA]: 50ka \citep{Fu2014-am,Hublin2020-nk}
%Add Mellars 2011 "The earliest modern humans in Europe", Nature. This is a nice commentary and Mellars was a wel-respected synthesizer of the European UP. Hajdinjak is not a paleo citation and not direct.
    \item Neanderthal introgression to Eurasian branch: 45ka \citep{Reilly2022-ym}
    \item Mende population expansion: 5ka
%Add Tennessen et al. 2012 "Evolution and Functional Impact of Rare Coding Variation from Deep Sequencing of Human Exomes"
    \item Gumuz population decline: 5ka
    %We actually don't have a citation for this. Leaving blank. 
    \item East African pastoralist to South African admixture: 2ka 
    \item European to South African admixture: 10 generations ago \citep{Uren2016-nn}
    \item South African population decline following colonial admixture: 9 generations ago \citep{Van_Eeden2022-od}
\end{itemize}
The total number of parameters that were ultimately inferred were 21-31, depending on the complexity
of the model.

@@

In our preliminary models, the unconstrained inferred split times between Eastern
African populations and Europeans was inferred to have occurred around $40$ka,
and the inferred split between Eastern and Western African populations occurred about
$10$ka earlier. These recent split dates are at odds with paleoanthropological 
evidence for an earlier Out-of-Africa (OOA) expansion
\citep[e.g.][]{Fu2014-am,Hublin2020-nk} and with the estimated date
for Neanderthal admixture into Eurasian populations
\citep[e.g.][]{Reilly2022-ym}. The parameter
uncertainty of our inferred divergence times (which accounts for the
non-independence of sites) does not account for these differences: in the
high-likelihood models, the $95\%$ confidence intervals for the Eastern
African-European split was roughly $38 \pm 5$ka, and for the Eastern/Western African
split was $45 \pm 10$ka, with the uncertainties reflecting 2 standard
deviations according to bootstrap of genomic regions. We therefore expect these
discrepancies to result from model misspecification in the details of the
divergence between the ancestors of OOA and Eastern African populations.

@@

In our primary model fits, we used the Mende in Sierra Leone (MSL) and the
British from England and Scotland (GBR). We tested the same models against a
different West African population (the Yoruba from Ibidan, Nigeria, YRI), and
two different European populations (Western Europeans from Utah, CEU, and
Toscani in Italy, TSI). The CEU are closely related to GBR and are thus likely
to be an equivalently close proxy for European ancestry in African individuals
with admixed European ancestry. The TSI are likely more distantly related to
the source of this European ancestry. We chose not to replace the GBR
by non-European Eurasian populations, as these would be poorer proxies for early
Holocene and recent colonial migration from the Near East and Europe back to
Africa, and it is unclear how such a model should be interpreted. We therefore
refit each of our best-fit demographic models using the following alternate
population sets: (Nama, YRI, Gumuz, Oromo/Amhara, and CEU), and (Nama, MSL,
Gumuz, Oromo/Amhara, and TSI).

@@

\subsubsection{Replacing Thousand Genomes populations}
\label{sec:replacing}
We replaced the Mende (MSL) and British (GBR) population sets with Yoruba (YRI)
and Western and Northern Europeans (CEU), respectively, and refit our inferred
demographic models to diversity and LD statistics computed from those
populations (along with Nama, Oromo/Amhara, and Gumuz). Because the MSL and YRI
share recent West African ancestry, and the GBR and CEU share recent European
ancestry, we expected most parameters in inferred models using these two
datasets to be largely similar. Indeed, all parameters inferred using the YRI
and CEU data were well within one standard error of the same parameters
inferred using the MSL and GBR data, across all tested models. The one
exception to this was the recent size of the YRI branch, inferred to be
somewhat smaller than the final size of the analogous MSL branch ($\approx
20,000$ vs. $\approx 30,000$, resp.). Other population sizes, migration rates,
and divergence times were highly consistent.

@@

In the primary results presented in the main text
(Tables~\ref{tab:supp-single-origin}--\ref{tab:supp-merger-with-stem-migration}),
we fixed a number of parameters that have been previously inferred or are well
constrained by historical records (Section~\ref{sec:models}), including the
Eurasian out-of-Africa expansion at 50ka and the divergence of Western and
Eastern African groups at 60ka. To evaluate the effects of these choices on the
inference of other parameters in the model, we first fixed those modeled
divergence times to 60ka (for the out-of-Africa expansion) and 100ka (for the
E/W African divergence). In the next section, we allow those parameters to be
fit along with all other model parameters.

@@

When we allowed the out-of-Africa expansion and Eastern/Western African
divergence times to be jointly fit in model optimization, we found that the
best-fit inferred E/W divergence time and the OOA expansion were more recent
($45-55$ka and $35-40$ka, resp.). The remainder of the parameters were largely
unaffected by these shifts to more recent divergence times, and our
interpretation of the inferred early history remains unchanged. The $95\%$
confidence intervals around these parameters were relatively narrow ($\approx
5$ka for the OOA split, $\approx 10$ka for the East/West African split),
although this uncertainty does not capture uncertainty due to model
mis-specification (as discussed in Section~\ref{sec:models}).

>>>>

Our models also included recent events to account for known migrations,
admixtures, and growths and declines in effective population sizes. Many of
these parameters were fixed based on previous historical, genetic, or
anthropological research, namely
\begin{itemize}
    \item Western African vs. eastern African divergence: 60ka
        \citep{Pagani2015-pz,Chevrier2018-vv,Niang2020-xt}
    \item Eastern Africa vs. British divergence (i.e., OOA): 50ka
        \citep{Mellars2011-sr,Fu2014-am,Hublin2020-nk}
    \item Neanderthal introgression to Eurasian branch: 45ka \citep{Reilly2022-ym}
    \item Mende population expansion: 5ka \citep{Tennessen2012-oy}
    \item Gumuz population decline: 5ka
    %We actually don't have a citation for this. Leaving blank. 
    \item East African pastoralist to South African admixture: 2ka \citep{Henn2008-xo}
    \item European to South African admixture: 10 generations ago \citep{Uren2016-nn}
    \item South African population decline following colonial admixture: 9 generations ago \citep{Van_Eeden2022-od}
\end{itemize}
The total number of parameters that were ultimately inferred were 21-31, depending on the complexity
of the model.

@@

In our preliminary models, the unconstrained inferred split times between eastern
African populations and Europeans was inferred to have occurred around $40$ka,
and the inferred split between eastern and western African populations occurred about
$10$ka earlier. These recent split dates are at odds with paleoanthropological 
evidence for an earlier Out-of-Africa (OOA) expansion
\citep[e.g.][]{Fu2014-am,Hublin2020-nk} and with the estimated date
for Neanderthal admixture into Eurasian populations
\citep[e.g.][]{Reilly2022-ym}. The parameter
uncertainty of our inferred divergence times (which accounts for the
non-independence of sites) does not account for these differences: in the
high-likelihood models, the $95\%$ confidence intervals for the eastern
African-European split was roughly $38 \pm 5$ka, and for the eastern/western African
split was $45 \pm 10$ka, with the uncertainties reflecting 2 standard
deviations according to bootstrap of genomic regions. We therefore expect these
discrepancies to result from model misspecification in the details of the
divergence between the ancestors of OOA and eastern African populations.

@@

In our primary model fits, we used the Mende in Sierra Leone (MSL) and the
British from England and Scotland (GBR). We tested the same models against a
different West African population (the Yoruba from Ibidan, Nigeria, YRI), and
two different European populations (western Europeans from Utah, CEU, and
Toscani in Italy, TSI). The CEU are closely related to GBR and are thus likely
to be an equivalently close proxy for European ancestry in African individuals
with admixed European ancestry. The TSI are likely more distantly related to
the source of this European ancestry. We chose not to replace the GBR
by non-European Eurasian populations, as these would be poorer proxies for early
Holocene and recent colonial migration from the Near East and Europe back to
Africa, and it is unclear how such a model should be interpreted. We therefore
refit each of our best-fit demographic models using the following alternate
population sets: (Nama, YRI, Gumuz, Oromo/Amhara, and CEU), and (Nama, MSL,
Gumuz, Oromo/Amhara, and TSI).

@@

\subsubsection{Replacing Thousand Genomes populations}
\label{sec:replacing}
We replaced the Mende (MSL) and British (GBR) population sets with Yoruba (YRI)
and western and northern Europeans (CEU), respectively, and refit our inferred
demographic models to diversity and LD statistics computed from those
populations (along with Nama, Oromo/Amhara, and Gumuz). Because the MSL and YRI
share recent West African ancestry, and the GBR and CEU share recent European
ancestry, we expected most parameters in inferred models using these two
datasets to be largely similar. Indeed, all parameters inferred using the YRI
and CEU data were well within one standard error of the same parameters
inferred using the MSL and GBR data, across all tested models. The one
exception to this was the recent size of the YRI branch, inferred to be
somewhat smaller than the final size of the analogous MSL branch ($\approx
20,000$ vs. $\approx 30,000$, resp.). Other population sizes, migration rates,
and divergence times were highly consistent.

@@

In the primary results presented in the main text
(Tables~\ref{tab:supp-single-origin}--\ref{tab:supp-merger-with-stem-migration}),
we fixed a number of parameters that have been previously inferred or are well
constrained by historical records (Section~\ref{sec:models}), including the
Eurasian out-of-Africa expansion at 50ka and the divergence of western and
eastern African groups at 60ka. To evaluate the effects of these choices on the
inference of other parameters in the model, we first fixed those modeled
divergence times to 60ka (for the out-of-Africa expansion) and 100ka (for the
E/W African divergence). In the next section, we allow those parameters to be
fit along with all other model parameters.

@@

When we allowed the out-of-Africa expansion and eastern/western African
divergence times to be jointly fit in model optimization, we found that the
best-fit inferred E/W divergence time and the OOA expansion were more recent
($45-55$ka and $35-40$ka, resp.). The remainder of the parameters were largely
unaffected by these shifts to more recent divergence times, and our
interpretation of the inferred early history remains unchanged. The $95\%$
confidence intervals around these parameters were relatively narrow ($\approx
5$ka for the OOA split, $\approx 10$ka for the East/West African split),
although this uncertainty does not capture uncertainty due to model
mis-specification (as discussed in Section~\ref{sec:models}).
